\subsection{A simplified ledger model}

We use the term \textit{ledger} to refer to the state of the blockchain. Entities on the ledger have addresses. The three kinds of addresses are \textit{account addresses}, \textit{package addresses} and \textit{cell addresses} (or just addresses for short).

Account addresses act as the identities of external actors. They allow them to interact with the blockchain, similarly to EOAs (Externally Owned Accounts) on Ethereum.

All Axi code deployed to the ledger is stored in a global codebase, divided into \textit{packages}. Package addresses are used to locate these packages in the codebase. Packages are further subdivided into smaller units, called \textit{modules}, which hold definitions of types, traits, constants and functions.

Data and resources on the ledger are stored in \textit{cells}. Each cell has an address and a type, and stores exactly one value of that type. Cells also have some additional metadata, which will not concern us here.

Cells are either \textit{owned} or \textit{shared}. Owned cells have exactly one owner, which is an account. Shared cells do not have an owner. Shared cells are further subdivided into \textit{mutable} shared cells and \textit{immutable} shared cells. Mutable shared cells contain resources and can be interacted with by anyone (any account or cell) according to the interface set by the package defining the cell's type. Immutable shared cells contain data (but not resources) and can be read by everyone, but cannot be modified by anyone.

The state of the ledger is modified by transactions. Every transaction has a \textit{transactor} (also called the \textit{transaction originator}) and a \textit{transaction body}. There are two types of transactions: \textit{code deploys} and \textit{ordinary transactions}. The body of a code deploy transaction is a package of Axi code. The body of an ordinary transaction is an Axi function that describes how to modify the ledger. Transactions also have some additional metadata, which will not concern us here.